On the propriety of restricted Boltzmann machines

### 2016-10-24 writing/draft.Rmd
notation on encoding cleared up

---
title: "On the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "September, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, and two layers, one hidden and one visible, with no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thereby thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has largely been unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model behaviors such as degeneracy, instability and/or uninterpretability. We also describe the difficulties that arise in likelihood-based fitting of such highly flexible models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
---

@@

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples]. A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, and with no conditional dependency within a layer [@smolensky1986information]. These models have reportedly been used with success in classification of images [@larochelle2008classification, @srivastava2012multimodal]. However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]. In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical theory and then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

First, the models are often unsatisfactory as a conceptualization of how data are generated. According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical]. Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination]. At issue, RBMs can often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of the data generation process. In addition to such degeneracy, we find that RBMs can easily exhibit types of instability in the parameter space. In practice, this may seen when a single pixel change in an image results in a wildly different classification. Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing, @nguyen2014deep]. We investigate and these phenomena for RBMs in Section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly being intractable. But even when fitting is feasible, there are unique challenges due to the overparameterization of RBMs. The literature often suggests Markov chain Monet Carlo (MCMC) approximations to fit these models (e.g., Gibbs sampling to exploit conditional structure in the hidden and visible variables), but little is said about monitoring convergence [@hinton2010practical, @hinton2006fast]. Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) per observation (visibles) which can critically influence convergence in MCMC-based likelihood methods. In Section \ref{bayesian-model-fitting}, we compare three Bayesian modeling techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models. As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of feasibility. Additionally, because RBM models are highly overparametrized, any fully principled fitting method based on the likelihood will simply reproduce the (discrete) empirical distribution of most realistic training sets. That is, there is no "smoothed distribution" achieved in fitting a RBM model with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.

This paper is structured as follows. Section \ref{background} formally definesthe RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}. Additionally, there is an accompanying appendix with proofs of main results.

@@

### Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_H, v_1,\dots,v_V)$ represent the states of the visible and hidden nodes in an RBM for some integers $V, H \ge 1$. We will consider both $\boldsymbol x \in \{0,1\}^{H + V}$ and $\boldsymbol x \in \{-1, 1\}^{H + V}$, as the potential encodings for a binary $H + V$-dimensional space, $\mathcal{X}$. We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_H, V_1,\dots,V_V)$, for the nodes as

@@

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathbb{R}^{V + H + V*H}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$ is the normalizing function to ensure the probabilities (\ref{eqn:pmf}) sums to one. For  
\begin{align}
\boldsymbol t(\boldsymbol x) = (h_1, \dots, h_H, v_1, \dots, v_V, v_1 h_1, \dots, v_V h_H) \in \{0,1\}^{H + V + H*V}, \boldsymbol x = (h_1, \dots, h_H, v_1, \dots, v_V) \in \mathcal{X} = \{0,1\}^{V + H}, \label{eqn:t}
\end{align}
let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$ be the vector of random variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ for the neg-potential function. Note that the RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, V,\\j = 1, \dots, H}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion. A kind of Gibbs sampling can be tried, which is natural due to conditional independence architecture within the RBM (i.e. visibles given hiddens or vice versa). Specifically, the conditional independence of nodes in each layer, given those nodes in the other layer allows for stepwise simulation of both the visible and hidden layers.

@@

RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples]. By stacking multiple layers of RBMs in a deep architecture, proponents of the models claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp. 450]. The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on until the desired multi-layer architecture is created.

## Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of a RBM (having as it does $H + V + H V$ parameters) creates three kinds of potential model impropriety of concern: *degeneracy, instability, and uninterpretability*. In this section we define these characteristics, consider how to detect them in a RBM, and point out their relationships to one another.

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph. Again, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function for a valid distribution. Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}t(X)$, is close to the boundary of $C$.

The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* as follows. When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H ) \in \boldsymbol \tau \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X} = \{0,1\}^{V + H}\} \subset \{0,1\}^{V + H + V*H}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\boldsymbol \tau$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is given as
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\}.
$$

@@

Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*. Instability can be roughly characterized by excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability), it is useful to imagine how the RBM model might be expanded to incorporate more and more observations. When increasing the size of a RBM model to handle more visible variables, it becomes necessary to expand the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables to be included). To this end, let $\boldsymbol \theta_{V} \equiv (\theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}, \theta_{11}, \dots, \theta_{VH}), V \ge 1$, denote a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ as
\begin{align}
\frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}\right] \equiv \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \label{eqn:elpr}
\end{align}
where $P_{\boldsymbol \theta_V}(v_1, \dots, v_V) \propto \sum_{\{h_j\}\in \{-1,1\}}\exp\left(\sum_{i = 1}^V \theta_{v_i} v_i + \sum_{j = 1}^H \theta_{h_j} h_j + \sum_{i = 1}^V \sum_{j = 1}^H \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_V)$ for the visible variables $(V_1, \dots, V_V)$ under parameters $\boldsymbol \theta_V$.

In formulating a RBM model for a potentially large number of visibles (i.e., $V \rightarrow \infty$), we will see that the ratio (\ref{eqn:elpr}) needs to stay bounded for the RBM model to be stable. This leads to the following definition:

\begin{defn}[S-unstable RBM]
Let $\boldsymbol \theta_V \in \mathbb{R}^{V + H +  H*V}, v \ge 1$, be a sequence of RBM parameters where the number of hiddens, $H \equiv H(V) \ge 1$, can be an arbitrary function of the number $V$ of visibles. A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if, as the number of visible variables increase ($V \rightarrow \infty$), it holds that 
\begin{align*}
\lim\limits_{V \rightarrow \infty} \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \lim\limits_{V \rightarrow \infty} \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $V_C > 0$ so that $\frac{1}{V}\text{ELPR}(\boldsymbol \theta_V) > C$ for all $V \ge V_C$.

Note that this definition of *S-unstable* is a generalization or re-interpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models for visibles $(v_1, \dots, v_V)$ and an increasing number of parameters.

S-unstable RBM models are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_V) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*)} : (v_1, \dots, v_V) \& (v_1^*, \dots, v_V^*) \in \{-1,1\}^V \text{ differ by exactly one component}\right\},
$$
denoting the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_V$. We then have the following result.

@@

In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then the RBM model will exhibit large probability shifts for very small changes in the data configuration (i.e., instability).

Additionally, S-unstable RBM models are connected to degenerate models  placing all probability on a small portion of the sample space. To see this, define a modal set of possible outcomes (i.e. set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_V, V \ge 1$ as
$$
M_{\epsilon, \boldsymbol \theta_V} \equiv \left\{(v_1, \dots, v_V) \in \{-1,1\}^V: \log P_{\boldsymbol \theta_V}(v_1, \dots, v_V) > (1-\epsilon)\max\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) + \epsilon\min\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) \right\}
$$
for a given $0 < \epsilon < 1$. Then S-unstable models are guaranteed to be degenerate, as the following result shows.

>>>>

---
title: "On the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "September, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thereby thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has been largely unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model behaviors such as degeneracy, instability and/or uninterpretability. We also describe the difficulties that arise in likelihood-based fitting and Bayes of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
---

@@

The data mining and machine learning communities have recently shown a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples]. A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information]. These models have reportedly been used with success in classification of images [@larochelle2008classification, @srivastava2012multimodal]. However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]. In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical theory, and then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

First, the models are often unsatisfactory as a conceptualization of how data are generated. According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical]. Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination]. At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of the data generation process. In addition to such degeneracy, we find that RBMs can easily exhibit types of instability. In practice, this may seen when a single pixel change in an image results in a wildly different classification in an image classification problem. Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing, @nguyen2014deep]. We investigate and these phenomena for RBMs in Section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable. The literature often suggests Markov chain Monte Carlo (MCMC) approximations to posteriors to fit these models (e.g., Gibbs sampling to exploit conditional structure in the hidden and visible variables), but little is said about acheiving convergence [@hinton2010practical, @hinton2006fast]. Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) per observation (visibles) which can critically influence convergence in MCMC-based likelihood methods.

In Section \ref{bayesian-model-fitting}, we compare three Bayesian modeling techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models. As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.

Additionally, because RBM models are highly overparametrized, any fully principled fitting method based on the likelihood will simply reproduce the (discrete) empirical distribution of most realistic training sets. That is, there is no "smoothed distribution" achieved in fitting a RBM model with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.

This paper is structured as follows. Section \ref{background} formally definesthe RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of detecting it are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques intended to avoid fitted model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}. Additionally, there is an accompanying appendix with proofs of main results.

@@

### Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_H, v_1,\dots,v_V)$ represent the states of the visible and hidden nodes in an RBM for some integers $V, H \ge 1$. Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$ with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural or na{\"i}ve starting point, we show later that the coding $\mathcal{C}=\{-1,1\}$ induces better model properties for the RBM. We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_H, V_1,\dots,V_V)$, for the nodes as

@@

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathbb{R}^{V + H + V*H}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{H+V}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one. For 
\begin{align}
\boldsymbol t(\boldsymbol x) = (h_1, \dots, h_H, v_1, \dots, v_V, v_1 h_1, \dots, v_V h_H) \in \mathcal{C}^{H + V + H*V}, \boldsymbol x = (h_1, \dots, h_H, v_1, \dots, v_V) \in \mathcal{C}^{V + H}, \label{eqn:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$ be the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ for the neg-potential function. Note that the RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, V,\\j = 1, \dots, H}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion. A kind of Gibbs sampling can be tried, which is natural due to conditional independence architecture of the RBM (i.e. visibles given hiddens or vice versa). Specifically, the conditional independence of nodes in each layer, given those nodes in the other layer allows for stepwise simulation of both the visible and hidden layers.

@@

RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples]. By stacking multiple layers of RBMs in a deep architecture, proponents of the models claim to produce the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp. 450]. The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on until the desired multi-layer architecture is created.

## Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of a RBM (having as it does $H + V + H V$ parameters) creates at least three kinds of potential model impropriety of concern that we will call *degeneracy, instability, and uninterpretability*. In this section we define these characteristics, consider how to detect them in a RBM, and point out relationships among them.

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph. Again, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t(X)$, is close to the boundary of $C$.

The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* as follows. When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{H + V} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\}.
$$

@@

Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*. Instability can be roughly characterized as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how the RBM model might be expanded to incorporate more and more visibles When increasing the size of a RBM model, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{V} \equiv (\theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}, \theta_{11}, \dots, \theta_{VH}), V \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ as
\begin{align}
\frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}\right] \equiv \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \label{eqn:elpr}
\end{align}
where $P_{\boldsymbol \theta_V}(v_1, \dots, v_V) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^V \theta_{v_i} v_i + \sum_{j = 1}^H \theta_{h_j} h_j + \sum_{i = 1}^V \sum_{j = 1}^H \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_V)$ for the visible variables $(V_1, \dots, V_V)$ under parameters $\boldsymbol \theta_V$.

In formulating a RBM model for a potentially large number of visibles (i.e., $V \rightarrow \infty$), we will say that the ratio (\ref{eqn:elpr}) needs to stay bounded for the sequence of RBM models to be stable. That is, we make the following convention

\begin{defn}[S-unstable RBM]
Let $\boldsymbol \theta_V \in \mathbb{R}^{V + H +  H*V}, v \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $H \equiv H(V) \ge 1$, can be an arbitrary function of the number $V$ of visibles. A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{V \rightarrow \infty} \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \lim\limits_{V \rightarrow \infty} \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $V_C > 0$ so that $\frac{1}{V}\text{ELPR}(\boldsymbol \theta_V) > C$ for all $V \ge V_C$.

This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_V)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_V$ is not fixed, but rather grows with $V$.

S-unstable RBM model sequences are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_V) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*)} : (v_1, \dots, v_V) \& (v_1^*, \dots, v_V^*) \in \mathcal{C}^V \text{ differ by exactly one component}\right\},
$$
denoting the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_V$. We then have the following result.

@@

In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., instability).

Additionally, S-unstable RBM model sequences are connected to degenerate model sequences placing all probability on a small portion of their sample spaces. To see this, define a modal set of possible outcomes (i.e. set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_V, V \ge 1$ as
$$
M_{\epsilon, \boldsymbol \theta_V} \equiv \left\{(v_1, \dots, v_V) \in \mathcal{C}^V: \log P_{\boldsymbol \theta_V}(v_1, \dots, v_V) > (1-\epsilon)\max\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) + \epsilon\min\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) \right\}
$$
for a given $0 < \epsilon < 1$. Then S-unstable model sequences are guaranteed to be degenerate, as the following result shows.
